---
title: "Lab 06 - Dataset Joins"
author: "Kenaniah Dodson"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
    theme: cerulean
    highlight: haddock
    toc: yes
    toc_float: yes
---

```{r include = FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.width = 10)     # Global knitr options

library(dplyr)
library(pander)
library(Lahman)                           # Load packages

data(Teams)
data(Salaries)                            # Load data

options(scipen = 999)                     # Disable sci. notation

```

<br>

# Introduction

The following work identifies the baseball teams that have achieved the largest 
number of wins on the smallest budgets since 1985 per the `Lahman` data package.

<br>

# Preprocessing

This analysis links two `Lahman` package tables, viz. `Salaries` and `Teams`, 
to determine total wins per season and the salary cost of each win earned. See 
the following preview of the first observations in `Salaries`:

<br>

```{r echo = FALSE}

head(Salaries) %>%
  pander()

```

<br>

See the following preview of select variables and first observations in `Teams`:

<br>

```{r echo = FALSE}

Teams %>%
  select(yearID, teamID, lgID, G, W, L) %>%
  head() %>%
  pander()

```

<br>

Lastly, `Salaries` data are adjusted for inflation at a constant annual increase 
rate of 3%. Observe:

<br>

```{r echo = TRUE}

Salaries <- Salaries %>% 
  mutate(salary.adj = salary * (1.03)^(max(yearID) - yearID))

head(Salaries) %>%
  pander()

```

<br>

In order to calculate the cost per win, adjusted individual player salaries from 
the `Salaries` table must be aggregated by team and year in the `Teams` table.

<br>
<br>

# Step 1: Aggregating Salaries

Using the `Salaries` table and grouping on variables `yearID` and `teamID` via
package `dplyr` function `group_by()`, the following uses function 
`summarize()` to *calculate and name* the following summary variables for 
*each year and team*: 

<br>

* Total adjusted team budget using `salary.adj` and `sum()`
 
* Total players using `n()`

<br>

```{r}

# Your code here
  

  Salaries_summary <- Salaries %>%
  group_by(yearID, teamID) %>%
  summarise(team.budget = sum(salary.adj), count = n())


Salaries_summary %>% head() %>% pander()
  




```

<br>
<br>

# Step 2: Join Salaries & Teams

The following merges the `Salaries` summary data and full `Teams` tables using 
either base R function `merge()` or package `dplyr` functions `left_join()`. 

<br>

* Variables `teamID` and `yearID` are unique merging keys
 
* Example: `dataset_1 %>% left_join(dataset_2)`

<br>

```{r}

# Your code here

Teams_selected <- Teams %>%
  select(yearID, teamID, lgID, G, W, L)

merged_data <- left_join(Teams_selected, Salaries_summary, 
                         by = c("yearID", "teamID"))

merged_data %>% head() %>% pander()





```

<br>
<br>

# Step 3: Calculate Cost per Win

The following uses the merged `Salaries` summary data and `Teams` data to 
create a new variable with function `mutate()` that calculates the total 
team salary cost per win, divided by $100,000.

<br>

* Divide (`/`) total salaries from **Step 1** by total wins (`W`)
 
* Divide (`/`) that value by `100000`

<br>

```{r}

# Your code here

merged_data <- merged_data %>%
  mutate(cost.per.win = (team.budget / W) / 100000)

merged_data %>% head() %>% pander()





```

<br>
<br>

# Step 4: Select, Filter, & Arrange

The following filters out teams in merged data with less than 25 players on 
their annual roster. It then arranges teams in ascending order by cost per win. 
Lastly, only variables are preserved to create the full version of the 
table preview.

<br>

* Use `filter()` to remove observations with less than 25 players
 
* Use `arrange()` to order observations by total cost per win
 
* Use `select()` to preserve only essential variables

<br>

**Objective:** The following contains variables and initial values 
of the full table.  Variable names may differ.

<br>

```{r echo = FALSE}

data.frame(yearID = c(1988, 1986, 1987),
           teamID = c("PIT", "TEX", "ML4"),
           lgID = c("NL", "AL", "AL"),
           Rank = c(2, 2, 3),
           G = c(160, 162, 162),
           W = c(85, 87, 91),
           n = c(27, 26, 28),
           team.budget = c(13724134, 16367320, 17186960),
           cost.per.win = c(1.614604, 1.881301, 1.888677)) %>%
  pander()

```

<br>

**Solution:** Create the full version of the table below.

```{r}

# Your code here

final_table <- merged_data %>%
  filter(count >= 25) %>%
  arrange(cost.per.win) %>%
  select(yearID, teamID, lgID, G, W, count, team.budget, cost.per.win)

final_table %>% head() %>% pander()


```

<br>
<br>

